The value of being indispensable: a cooperative approach to bargaining with a continuum of players

### 2023-11-22 src/paper/theory.tex
Remove example application

\section{Example Application}

This section demonstrates the proposed cooperative approach via a model of a vertical market.
Note, that this is not a full-fledged model, and many factors are exogenous.
However, it is sufficient to illustrate the advantages of the framework presented in this paper: most notably, its analytical tractability and that it captures what intuitively feels like bargaining power.

Consider a market in the vein of \textcite{hart1990property}, but with a finite number of upstream and a continuum of downstream firms.
Furthermore, let us model production and sales in a reduced form, without going into the micro-foundations of the profit functions.
Finally, let us assume that profits are shared according to players' Shapley-values.

In the rest of this section, I will examine the effect of various market structures on the profit distribution.
I start by looking at the case of a single upstream producer, with no direct acces to the downstream market.
I then contrast this with two scenarios in which the upstream producer acquires a way to directly sell to the final consumers: either by vertical integration (merging with some share of the downstream firms), or by creating a completely new, downstream division within itself.

As this model is only supposed to be an illustration, I will focus on simplicity rather than realism.
Furthermore, I will not endogenize certain choices, such as entry, or the actual merger decisions.
Nevertheless, some of the results are suggestive of what might happen in an applied, more fully specified model.

\subsection{Benchmark case}

Let us start by looking at the simplest case: one upstream producer, no direct access to the downstream market.
Assume that without the upstream producer, the downstream firms cannot produce anything, and thus total industry profits are zero.
On the other hand, if the upstream producer and $t$ fraction of the downstream firms create an agreement, then total industry profits are given by the function $f(t)$.\footnote{
    This reduced-form profit function can be given microfoundations in many ways.
    For an example, see \textcite{stancsics2023hybrid}.
}
Furthermore, the following assumption guarantees ensures that the game is superadditive, therefore it actually makes sense to have every downstream firm in the coalition.\footnote{
    TODO: elaborate on what this assumption entails.
}
\begin{assumption}
    Let us assume that $f$ is continuous and increasing in $t$.
\end{assumption}

As noted before, profits are distributed according to players' Shapley values.
Consequently, it follows from \cref{prop:one_sided} that the profits of the upstream producer and the downstream mass of firms are
\begin{align*}
    \pi_P^b & = \int_0^1 f(t) \dt, \\
    \pi_A^b & = f(1) - \int_0^1 f(t) \dt,
\end{align*}
respectively.

\subsection{Vertical integration}

Let us now consider the case in which the upstream producer merges with some fraction $\alpha$ of the downstream firms.
The total profits are the same as before, namely, $f(1)$.\footnote{
    TODO: do I need a footnote here?
}
However, the profit distribution is now different.
New shares can be calculated as follows:
\begin{align}
    \pi_{P}^m & = \frac{1}{1 - \alpha} \int_\alpha^1 f(t) \dt, \\
    \pi_{A}^m & = f(1) - \frac{1}{1 - \alpha} \int_\alpha^1 f(t) \dt .
\end{align}

Clearly, the upstream producer gets a larger share than before, while the remaining downstream firms capture a smaller share.
This is intuitive, as the upstream producer now depends on the remaining downstream sellers to a lesser extent.
However, it's worth examining whether this increase is large enough in the sense that the now hybrid upstream producer's profit is higher than the sum of its profit and $\alpha$ fraction of the downstream firms' profit from the benchmark case.
The participants only have an incentive to merge if this is the case.
It turns out that this depends on the shape of the profit function.\footnote{
    This way of modeling vertical integration assumes that the upstream firm's innate bargaining power is not affected by the merger.
    In the terms of some non-cooperative bargaining model, such as \textcite{hart1996bargaining}, it means that the non-acquired downstream firms get a higher probability of proposing allocations.
    It could be interesting to see how the results change when the upstream firm's also obtains the acquired downstream firms' innate bargaining power, and profits are distributed according to weighted Shapley values.
}

\begin{proposition}
    \label{prop:vertical_integration}
    If $f$ is concave (convex) on $[0, 1]$, then $\pi_P^m \geq \pi_P^b + \alpha \pi_A^b$ ($\pi_P^m \leq \pi_P^b + \alpha \pi_A^b$). 
\end{proposition}

\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \begin{tikzpicture}[scale=0.8]
            \begin{axis}[xmin=0, xmax=1, ymin=0, ymax=1, samples at={0, 0.02, ..., 0.98, 1},
                xtick={0, 1}, ytick={0, 1}, yticklabels={0, f(1)},
                axis lines=left, xlabel={$s$}]
                \addplot[name path=f, dashed, black] {sqrt(x)};
                \addplot[name path=tildef, black, ultra thick] {sqrt(0.3+0.7*x)};
                \path[name path=bottom] (axis cs:0,0) -- (axis cs:1,0);
                \draw[name path=top] (axis cs:0,1) -- (axis cs:1,1);
                \draw[name path=right] (axis cs:1,0) -- (axis cs:1,1);

                \addplot [fill=blue, fill opacity=0.05] fill between [of=tildef and bottom];
                \addplot [fill=red, fill opacity=0.05] fill between [of=tildef and top];

                \node[anchor=north west] at (axis cs: .05, .95) {Downstream};
                \node[anchor=south east] at (axis cs: .95, .05) {Upstream};
            \end{axis}
        \end{tikzpicture}
        \caption{Concave case}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \begin{tikzpicture}[scale=0.8]
            \begin{axis}[xmin=0, xmax=1, ymin=0, ymax=1, samples at={0, 0.02, ..., 0.98, 1},
                xtick={0, 1}, ytick={0, 1}, yticklabels={0, f(1)},
                axis lines=left, xlabel={$s$}]
                \addplot[name path=f, dashed, black] {(x)^1.5};
                \addplot[name path=tildef, black, ultra thick] {(0.3+0.7*x)^1.5};
                \path[name path=bottom] (axis cs:0,0) -- (axis cs:1,0);
                \draw[name path=top] (axis cs:0,1) -- (axis cs:1,1);
                \draw[name path=right] (axis cs:1,0) -- (axis cs:1,1);

                \addplot [fill=blue, fill opacity=0.05] fill between [of=tildef and bottom];
                \addplot [fill=red, fill opacity=0.05] fill between [of=tildef and top];

                \node[anchor=north west] at (axis cs: .05, .95) {Downstream};
                \node[anchor=south east] at (axis cs: .95, .05) {Upstream};
            \end{axis}
        \end{tikzpicture}
        \caption{Convex case}
    \end{subfigure}
    \caption{Illustration of \cref{prop:vertical_integration}}
    \label{fig:vertical_intergration}
\end{figure}

The intuition is that when $f$ is concave, then obtaining some fraction of the firms lets it create a (comparatively) large share of the total possible profits by itself, and its bargaining position is considerably strengthened.
In comparison, when $f$ is convex, then the upstream producer still has a large need for the rest of the downstream firms, even when it acquires some of them.
The bottom line is that industries which are characterized by the downstream firms being relatively substitutable are more susceptible to consolidation.

\subsection{Upstream firm builds downstream capacity}

Now let us consider another way of getting direct downstream access: instead of acquiring some of the existing downstream firms, the upstream producer creates new downstream capacity from scratch.
In contrast to the previous case, the total size of the pie also changes.
Let's assume that the upstream firm's downstream capacity is equal to that of $\alpha$ measure of the downstream firms.
Then, the total profits when everyone agrees to participate is given by $f(1 + \alpha)$, which is larger than $f(1)$ before.

As before, let us examine the distribution of the profits in this case as well:
\begin{align*}
    \pi_{P}^c & = \int_0^1 f(t + \alpha) \dt, \\
    \pi_{A}^c & = f(1 + \alpha) - \int_0^1 f(t + \alpha) \dt .
\end{align*}
Similarly to the vertical merger case, the now hybrid upstream producer's profits increase with the size of its downstream capacity.
However, it is not obvious how the downstream's profit changes.
The next proposition shows that it, again, depends on the shape of the profit function.
\begin{proposition}
    \label{prop:downstream_capacity}
    If $f$ is concave (convex) on $[0, 1+\alpha]$, then $\pi_A^m \geq \pi_A^b$ ($\pi_A^m \leq \pi_A^b$). 
\end{proposition}

\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \begin{tikzpicture}[scale=0.8]
            \begin{axis}[xmin=0, xmax=1, ymin=0, ymax=1.14, samples at={0, 0.02, ..., 0.98, 1},
                xtick={0, 1}, ytick={0, 1, 1.14}, yticklabels={$0$, $f(1)$, $f(1 + \alpha)$},
                axis lines=left, xlabel={$s$}]
                \addplot[name path=f, dashed, black] {sqrt(x)};
                \addplot[name path=tildef, black, ultra thick] {sqrt(0.3+x)};
                \draw[dashed] (axis cs:0,1) -- (axis cs:1,1);
                \path[name path=bottom] (axis cs:0,0) -- (axis cs:1,0);
                \draw[name path=top] (axis cs:0,1.14) -- (axis cs:1,1.14);
                \draw[name path=right] (axis cs:1,0) -- (axis cs:1,1.14);

                \addplot [fill=blue, fill opacity=0.05] fill between [of=tildef and bottom];
                \addplot [fill=red, fill opacity=0.05] fill between [of=tildef and top];

                \node[anchor=north west] at (axis cs: .05, 1.09) {Downstream};
                \node[anchor=south east] at (axis cs: .95, .05) {Upstream};
            \end{axis}
        \end{tikzpicture}
        \caption{Concave case}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \begin{tikzpicture}[scale=0.8]
            \begin{axis}[xmin=0, xmax=1, ymin=0, ymax=1.482, samples at={0, 0.02, ..., 0.98, 1},
                xtick={0, 1}, ytick={0, 1, 1.482}, yticklabels={$0$, $f(1)$, $f(1 + \alpha)$},
                axis lines=left, xlabel={$s$}]
                \addplot[name path=f, dashed, black] {(x)^1.5};
                \addplot[name path=tildef, black, ultra thick] {(0.3+x)^1.5};
                \draw[dashed] (axis cs:0,1) -- (axis cs:1,1);
                \path[name path=bottom] (axis cs:0,0) -- (axis cs:1,0);
                \draw[name path=top] (axis cs:0,1.482) -- (axis cs:1,1.482);
                \draw[name path=right] (axis cs:1,0) -- (axis cs:1,1.482);

                \addplot [fill=blue, fill opacity=0.05] fill between [of=tildef and bottom];
                \addplot [fill=red, fill opacity=0.05] fill between [of=tildef and top];

                \node[anchor=north west] at (axis cs: .05, 1.43) {Downstream};
                \node[anchor=south east] at (axis cs: .95, .05) {Upstream};
            \end{axis}
        \end{tikzpicture}
        \caption{Convex case}
    \end{subfigure}
    \caption{Illustration of \cref{prop:downstream_capacity}}
    \label{fig:upstream_capacity}
\end{figure}

The results are similar to the previous case.
When the downstream firms are more substitutable, the upstream firms' newly created capacity strongly reduces their bargaining power, and even though the total size of the pie increases, the downstream firms' share decreases not only in relative terms, but also in absolute terms.
In a model with endogenous entry and exit on the part of the downstream firms, this would most likely lead to a reduction in their number.
The implication then also resembles the one in the vertical merger case: industries when downstream firms are easily substitutable should be strongly scrutinized when an upstream firm seeks to build its own downstream capacity.

@@

The illustrative model also highlights the strengths of this approach.
Despite its simplicity, it highlights that the effects of vertical integration strongly depends on the substitutability of the downstream firms.
In a time when upstream firms and platforms are increasingly seeking to sell directly to the consumers\footnote{
    As some recent examples, consider the recent acquisition of Whole Foods by Amazon, Microsoft's ongoing efforts to buy Activision/Blizzard (a major video game publisher), or Tesla's or BMW's plans to sell cars directly without the help of dealerships.
}
, such results can be important for competition authorities.

There is a variety of ways through which this paper could be expanded or built upon.
On the theoretical side, a number of the assumptions, such as fringe agents achieving a zero payoff without the platform, could be relaxed,
Besides this, the convergence result for the weighted values could be strengthened if one could manage to show that any sequence along ever more refined partitions of the atomic players converges to the same limit \parencite[à la][]{fogelman1980asymptotic}.

>>>>

@@

% The illustrative model also highlights the strengths of this approach.
% Despite its simplicity, it highlights that the effects of vertical integration strongly depends on the substitutability of the downstream firms.
% In a time when upstream firms and platforms are increasingly seeking to sell directly to the consumers\footnote{
%     As some recent examples, consider the recent acquisition of Whole Foods by Amazon, Microsoft's ongoing efforts to buy Activision/Blizzard (a major video game publisher), or Tesla's or BMW's plans to sell cars directly without the help of dealerships.
% }
% , such results can be important for competition authorities.

There is a variety of ways in which this paper could be expanded or built upon.
On the theoretical side, a number of the assumptions, such as fringe agents achieving a zero payoff without the platform, could be relaxed,
Besides this, the convergence result for the weighted values could be strengthened if one could manage to show that any sequence along ever more refined partitions of the atomic players converges to the same limit \parencite[à la][]{fogelman1980asymptotic}.
